\bf Distortion corrected kernel density estimator on Riemannian manifolds

### 2024-09-08 jcgs/03-chap3-kderm.Rmd
Tas changes

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = T,
  messages = FALSE,
  warning = FALSE,
  eval = TRUE,
  include = TRUE
)
options(tinytex.verbose = TRUE)
library(tidyverse)
library(here)
library(patchwork)
library(kableExtra)
# Jmisc::sourceAll(here::here("R/sources"))
# set.seed(1234)
```

@@

Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Scott2015-vl and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic networks in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq]. For $m$-dimensional data where $m$ is even of moderate size, KDE is challenging due to the well-known curse of dimensionality. However, the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d\ll m$. The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da, and @Berenfeld2022-fd. However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice, with the exception of @berenfeld2021-den and @berenfeld2022-est which demonstrate the minimax optimality of certain types of KDE in the context of density estimation on manifolds. Our objective in this paper is to propose KDE for data that lie on some unknown manifold after embedding.

@@

While full details are provided in @Perrault-Joncas2013-pq, we briefly describe the Learn Metric algorithm here. There are four main steps in the algorithm. First, a weighted neighborhood graph is constructed, with edges between $\bm{p}_i$ and $\bm{p}_j$ when $\bm{p}_i$ is a fixed-radius nearest neighbor of $\bm{p}_j$ or vice versa, and edge weights depending on the distance between $\bm{p}_i$ and $\bm{p}_j$ on the manifold. The fixed-radius method considers all points within a specified radius of \(\sqrt{\epsilon}\), allowing the graph to adapt to varying densities in the data and to capture local structure more effectively. Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\sqrt{\varepsilon}$ is the radius parameter in the previous step and a constant value $c=0.25$ is used for the use of heat kernel in the weighted neighborhood graph. Third, a dimension reduction method is applied to obtain the output embedding $\bm{y}_1,\dots,\bm{y}_n$. Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator). Full details on these four steps are provided in \autoref{alg:learnmetric}. This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz] although our own results are based on a re-implementation of the algorithm in *R*. For computational efficiency, approximate nearest neighbor searching methods are implemented to construct the neighborhood graph [@Cheng2021-dh; @Perrault-Joncas2013-pq], which are then used in both steps 1 and 3. Two parameters, $c=0.25$ and $\sqrt{\varepsilon} = 0.4$, are set as suggested in the *megaman* library.

@@

The estimator has a similar structure to Equation \@ref(eq:denriem) with some key differences. To understand these differences, it is first critical to appreciate that the coordinates $\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length in Section 6.2 of @Perrault-Joncas2013-pq). This is crucial for two reasons. First, this implies that the term $\| \bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\|$ approximates the geodesic distance between $\bm{y}_i$ and $\bm{y}_j$. Second, the estimator in Equation \@ref(eq:denriem) is valid only when the coordinate mapping is the logarithmic map around $\bm{y}_i$, and it is this mapping that is approximated by $\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)$. For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation $\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)$ which ensures that the embedding approximates the logarithmic map. Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut].

One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\bm{H}_j$. To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead. We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density. The entire workflow is summarized in \autoref{fig:vkde}. The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\bm{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection. Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue. With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.

@@

There are several open questions to be explored. The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature. For densities on unknown manifolds, @Berenfeld2021-as also provide some insights on the assumptions of the kernel function in \@ref(eq:kernelcondition). The second one is that the proposed density estimation method is only limited to the embedding points. Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation and estimation on the boundary of manifolds [@Berry2017-rz] in the future.
The last question is related to the quality of the embedding from dimension reduction methods. Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediately clear. The embedding quality could be measured using one of the metrics discussed in the online supplementary material of @Cheng2021-dh, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way. The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers. However, the outperformance of DC-KDE over KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.

>>>>

---
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = T,
  messages = FALSE,
  warning = FALSE,
  eval = TRUE,
  include = TRUE
)
options(tinytex.verbose = TRUE)
library(tidyverse)
library(here)
library(patchwork)
library(kableExtra)
# Jmisc::sourceAll(here::here("R/sources"))
# set.seed(1234)
```

@@

Multivariate kernel density estimation (KDE see @Parzen1962-gt; @Scott2015-vl and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic networks in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac] and prime prediction via Twitter messages [@Gerber2014-tq]. For $m$-dimensional data where $m$ is even of moderate size, KDE is challenging due to the well-known curse of dimensionality. However, the manifold hypothesis suggests that many high-dimensional datasets belong to a $d$-dimensional manifold $M$ where $d\ll m$. The problem of KDE on manifolds has been well studied in @Hendriks1990-bt, @Izenman1991-cj, @Pelletier2005-vu, @Henry2013-da, and @Berenfeld2022-fd. However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice. The is only a small literature dealing with the case where the manifold is unknown. Recently, @berenfeld2021-den, @berenfeld2022-est and @divol2022measure demonstrated the minimax optimality of certain types of KDE on manifolds. In contrast, our objective in this paper is to propose a new KDE for data that lie on some unknown manifold found using embedding techniques.

@@

While full details are provided in @Perrault-Joncas2013-pq, we briefly describe the Learn Metric algorithm here. There are four main steps in the algorithm. First, a weighted neighborhood graph is constructed, with edges between $\bm{p}_i$ and $\bm{p}_j$ when $\bm{p}_i$ is a fixed-radius nearest neighbor of $\bm{p}_j$ or vice versa, and edge weights depending on the distance between $\bm{p}_i$ and $\bm{p}_j$ on the manifold. The fixed-radius method considers all points within a specified radius of \(\sqrt{\epsilon}\), allowing the graph to adapt to varying densities in the data and to capture local structure more effectively. As an alternative, K nearest neghbors could also be used. Second, the discrete Laplacian on this graph $\hat{\mathcal{L}}_{\varepsilon,n}$ is estimated [@Zhou2011-za], where $\sqrt{\varepsilon}$ is the radius parameter in the previous step and a constant value $c=0.25$ is used for the use of heat kernel in the weighted neighborhood graph. Third, a dimension reduction method is applied to obtain the output embedding $\bm{y}_1,\dots,\bm{y}_n$. Fourth, the Riemannian metric at each point is estimated by exploiting the connection between the Riemannian metric and the Laplace Beltrami operator (to which the graph Laplacian at Step 2 is a discrete estimator). Full details on these four steps are provided in \autoref{alg:learnmetric}. This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz] although our own results are based on a re-implementation of the algorithm in *R*. For computational efficiency, approximate nearest neighbor searching methods are implemented to construct the neighborhood graph [@Cheng2021-dh; @Perrault-Joncas2013-pq], which are then used in both steps 1 and 3. Two parameters, $c=0.25$ and $\sqrt{\varepsilon} = 0.4$, are set as suggested in the *megaman* library. Note that in \autoref{alg:learnmetric} $c$ and $\sqrt{\varepsilon} = 0.4$ enter multiplicatively therefore only one of these parameters requires tuning. We considered different values of $\sqrt{\varepsilon}$, noting that results were stable for a range of values close to the default value of $\sqrt{\varepsilon} = 0.4$.

@@

The estimator has a similar structure to Equation \@ref(eq:denriem) which can be seen as the counterpart to Equation \@ref(eq:denestimator) if the Riemannian is known exactly. There are however, some key differences between the two equations. To understand these differences, it is first critical to appreciate that the coordinates $\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)$ give an embedding that is approximately isometric in a small neighborhood around the $i^{th}$ observed point (this insight is discussed at length in Section 6.2 of @Perrault-Joncas2013-pq). This is crucial for two reasons. First, this implies that the term $\| \bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)\|$ approximates the geodesic distance between $\bm{y}_i$ and $\bm{y}_j$. Second, the estimator in Equation \@ref(eq:denriem) is valid only when the coordinate mapping is the logarithmic map around $\bm{y}_i$, and it is this mapping that is approximated by $\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)$. For this reason, there is a ratio of two determinants to ensure the density integrates to one, the first is a consequence of the mapping from the manifold to the coordinate system (from a dimension reduction algorithm), while the second is the transformation $\bm{H}^{-1/2}_i (\bm{y}_j - \bm{y}_i)$ which ensures that the embedding approximates the logarithmic map. Also worth noting is the resemblance between the estimator and multivariate variable bandwidth estimation [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut].

One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\bm{H}_j$. To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead. To account for distortion, this average should be weighted by the determinant of the Riemannian at each of the nearest neighbors. We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density. The entire workflow is summarized in \autoref{fig:vkde}. The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\bm{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection. Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue. With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.

@@

There are several open questions to be explored. The first involves the selection of bandwidth parameters and kernel types for kernel density estimation, which has been widely explored in many KDE-related literature. Our method requires the choice of a single scale $r$ that tunes the varying bandwidth matrix. In practice this could be selected using a grid search according to some criteria that depends on the downstream task. For instance for anomaly detection, this criterion may be a measure of the similarity of anomalies detected both in sample and out-of-sample found using cross validation. For densities on unknown manifolds, @Berenfeld2021-as also provide some insights on the assumptions of the kernel function in \@ref(eq:kernelcondition). The second one is that the proposed density estimation method is only limited to the embedding points. Although for the downstream task of anomaly detection, only density estimated on these points is needed, it would be worth exploring the out-of-sample density estimation and estimation on the boundary of manifolds [@Berry2017-rz] in the future.
The last question is related to the quality of the embedding from dimension reduction methods. Although our distortion correction method is fairly robust to different choices of manifold learning methods, in certain cases, when the data structure is too complex and the distortion is too severe to correct, the quantitative relationship between embedding quality and density estimation accuracy is not immediately clear. The embedding quality could be measured using one of the metrics discussed in the online supplementary material of @Cheng2021-dh, but when the ground truth densities are unknown, which is usually the case with real-world data set, it is hard to tell whether the distortion is corrected in the right way. The density estimates on the edges of the whole data structure could also be explored because most outer area points tend to be detected as outliers. However, the outperformance of DC-KDE over KDE has been shown in the higher dimensional simulation data and the electricity usage data, which are more related to real-life data sets.
